---
title: "Estimating population size and reduction from removals"
subtitle: "(with emphasis on wild deer and wild pigs)"
link-citations: true
author: "Peter Caley (peter.caley@csiro.au) with Andrew Bengsen and Dave Forysth"
date: "`r Sys.Date()`"
output:
  html_document: default
bibliography: references.bib
citation_package: natbib    
header-includes:
    - \usepackage[round]{natbib}
csl: wildlife-research.csl
---


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, 
               fig.width = 10, 
               fig.height = 5, 
               out.width="100%", 
               fig.align = "center", 
               comment="")
options(knitr.kable.NA = "", scipen = 22)
suppressPackageStartupMessages({
  library(tidyverse)
  library(rjags)
  library(here)
})
TEST <- FALSE
PLOT.IT <- TRUE
source("../Prep/functions.R")
```

# Introduction

Central to using culling/removal rate to estimate abundance is the underlying relationship between the rate of culling (predation) and population density/size --- the functional response. This is an example of an ``inverse problem'' in inference, where we seek to estimate the initial abundance/density on the basis of the rate of removal. 

Some key points to note. 

- If the functional response relationship is well known, then only a single shooting session covering a representative sample of the area of interest would be required to obtain an estimate of initial population size/density. Note there are a bunch of qualifiers to this. For example, if the density is above the level at at which shooting saturation occurs, then only the lower bound for the estimate of density will be estimated with any precision.  
- If the functional response relationship is not known, then it needs to be estimated, which requires that the proportion of the population removed during sessions is sufficient to generate a decline in the kill rate enabling calibration. If there is no detectable decline in the culling rate, then the estimated population under saturating forms of the functional response is infinite.
- Finally, there is likely to be prior information available on the likely density of animals within the area of interest, along with the dominant vegetation type and what effect this may have on shape of the functional response relationship. Where such information is available and can be expressed quantitatively, Bayesian statistical methods are typically the best way of incorporation. 

Examples of the application of predator-prey theory to the shooting of feral pigs from a helicopter include @Saunders&Bryant:1988, @Hone:1990PredPrey and @Choquenot.etal:1999. A more recent application to controlling wild deer (and a review of ungulates in general) is that of @Bengsen.etal:2023. These studies use a range of functions to describe the relationship between the number of animals shot per unit shooting effort and abundance/density. They are reviewed here.  

## Functional response types

### Ivlev 

@Bengsen.etal:2023 used a modified form of the Ivlev function to model the kill rate ($c$) achieved by a helicopter shooting team as a function of deer density ($x$). The function takes the form: 

$$
 c = -b + a(1 - e^{-dx}).
$$
Note that the parameters $a$ and $b$ aren't interpretable exactly as stated by @Bengsen.etal:2023. In particular, the density below which animals are effectively unable to be shot is not $b$, but rather the x-intercept (where the curve crosses the x axis) that is controlled by $b$ interacting with $d$. In a similar vein, the maximum rate of removal is $c_{max}=a-b$ (not to simply $a$ as stated). It is possible, however, to calculate that values for $a$ and $b$ that give the desired value for the density of animals unable to be removed ($x_{min}$). Substituting $a = c_{max} + b$ into the Ivlev function with $c$ set to zero enables the required value of $b$ to be estimated as:

$$
 b = c_{max}(e^{d x_{min}}-1).
$$
That said, examining the Ivlev functional response models fitted to the aerial shooting data of deer in @Bengsen.etal:2023 doesn't indicate that a portion of the population is unable to be shot (see their Fig. 4). The requirement to have a functional response where a portion of the population is unable to be shot arose during aerial shooting of wild pigs where the habitat included areas of dense vegetation @Choquenot.etal:1999. Modifying the Ivlev function to be non-zero is essentially changing it to a Type-III (i.e. sigmoidal) functional response.   

```{r, fig.cap="Example of Ivlev functional response (modified to only take positive values) with parameters chosen to achieve a maximum kill rate of 40 hr$^{-1}$ and a density of 2 below which culling becomes ineffective.", echo=!TRUE}
if(PLOT.IT) {
   a.hat <- ab.est(c.max=40, d=0.5, x.crit=2)[1]
   b.hat <- ab.est(c.max=40, d=0.5, x.crit=2)[2]
   pars <- list(a=a.hat, b=b.hat, d=0.5, x.vals=seq(0, 20, 0.1))
   with(pars, {
      Ivlev.pred <-Ivlev(x=x.vals, a=a, b=b, d=d, no.neg=TRUE) 
      plot(x.vals, Ivlev.pred , type='l', lwd=2,
      xlab="Prey density",  ylab="Kill rate",
	    ylim=c(min(Ivlev.pred)-1, 40), axes=F)
      axis(1, lwd=3, pos=-0.2)
      axis(2, lwd=3, pos=-0.25, las=2)
      abline(h=40, lty=2)
      legend(10, 30, paste0("a=", round(a,1), ", b=", round(b,1), ", d=", d), lty=(1), lwd=2, title="Parameter settings")
   })
}
```

The Ivlev function is particularly suited to the analysis of numerical response (e.g. @Caughley:1987Relationships) where the rate of population increase ($r$) can take on negative values, and the parameters can be sensibly interpreted. For example, $b$ is the maximum rate of decrease in the absence of food, and $r=a-b$ is the maximum rate of increase when food isn't limiting. It is, however, probably less well suited as a functional response, where the predation rate can't be negative. 


### Choquenot *et al* 1990

@Choquenot.etal:1999 used what they termed a ``consonant'' model used by @Short:1987FoodIntake to describe variation in the kill rate (kills h$^{-1}$) of a shooting team consisting of a pilot and single shooter as a function of pig density (pigs km$^{-2}$) as:

$$
 \text{kill rate} = a (1 - e^{-d(x-b)}).
$$

Where:\
$x=$ density of target/prey species (here wild pigs),\
$a=$ the maximum offtake rate (realised at large $x$),\
$b=$ density of target animals below which detection and killing in impossible, and\
$d=$ parameter controlling the rate of encounter between the hunting process and the target animal [ToDo. Check by simulation]

The parameterisation of this model as some advantages over the Ivlev one in terms of interpretability --- the maximum rate of shooting at saturation is specified directly by $a$ (rather than $a - b$), and the density at which animals cannot no longer be detected is exactly $b$ (c.f. being calculated numerically). This is an advantage when it comes to using prior knowledge within a Bayesian modelling framework, in that the specification of prior distributions is straightforward --- not a consideration when the paper was being written. 

Note that the units of the kill rate could also be expressed as kills per area per unit time. 

```{r, fig.cap="Example of functional response used by @Choquenot.etal:1999.", include=TRUE, echo=!TRUE}
if(PLOT.IT) {
  pars <- list(a=40, b=2, d=0.5, x.vals=seq(0, 20, 0.1))
  with(pars, {
	    plot(x.vals, Choquenot(x.vals,a,b,d), type='l', lwd=2, xlim=c(0, max(x.vals)), 
	         xlab="Prey density", ylab="Kill rate", axes=F)
      axis(1, lwd=3, pos=-0.2)
      axis(2, lwd=3, pos=-0.25, las=2)
      abline(h=40, lty=2)
       legend(10, 30, paste0("a=", a, ", b=", b, ", d=", d), lty=(1), lwd=2, title="Parameter settings")
  })
}
```

The specification of a proportion of animals being unable to be preyed on is an approximation (e.g. where a proportion of animals learn to use and have access to refuges with zero visibility). It's inclusion by @Choquenot.etal:1999 presumably reflects the first-hand experience of the authors. For example, @Saunders&Bryant:1988 document pigs that were known (from radio-collaring data) to be present though never sighted during a helicopter shooting exercise due to hiding in tall reeds.  It assumes that removing animals becomes infinitely expensive (by shooting from a helicopter at least). 


### Alternatives Type III functional responses

It is somewhat strange that @Choquenot.etal:1999 didn't choose one of the standard functional responses relating predation rate to prey density. For example, following @Pech.etal:1992 a Type III response can be of the form:

$$
   c = \frac{\beta_0 x^2}{\beta_1^2 + x^2}
$$

Here the kill rate saturates at $\beta_0$. A limitation of this 2-parameter form of a Type III response is that it is difficult to have a pronounced sigmoidal shape that captures a significant density of prey that are unavailable, without a pronounced flattening of the curve (i.e. the shooting rate is slow to reach saturation). This can be addressed by allowing the exponent of $x$ to be free ($k$). Increasing $k$ shortens the time to reaching shooting saturation. Combinations of high values of both $k$ and $\beta_1$ can produce more pronounced sigmoidal shapes, similar to the 'generalised' Type III functional response (see below). 


```{r, fig.cap="Example of Holling Type III functional response used by @Pech.etal:1992.", include=TRUE, echo=!TRUE}
if(PLOT.IT) {
  pars1 <- list(beta0=40, beta1=2, k=2, x.vals=seq(0, 20, 0.1))
  pars2 <- list(beta0=40, beta1=2, k=3, x.vals=seq(0, 20, 0.1))
  pars3 <- list(beta0=40, beta1=20, k=4, x.vals=seq(0, 20, 0.1))
  with(pars1, {
	     plot(x.vals, TypeIII(x.vals, beta0, beta1), type='l', lwd=2, xlim=c(0, max(x.vals)), xlab="Prey density", ylab="Kill rate", axes=F)
      axis(1, lwd=3, pos=-0.2)
      axis(2, lwd=3, pos=-0.25, las=2)
      abline(h=40, lty=2)
      legend(10, 30, paste0("beta0=", beta0, ", beta1=", beta1, ", k=", k), lty=c(1), lwd=2, title="Parameter settings", bty='n')
  })
  with(pars2, {
	     lines(x.vals, TypeIII(x.vals, beta0, beta1, k), lty=2, lwd=2)
       legend(10, 20, paste0("beta0=", beta0, ", beta1=", beta1, ", k=", k), lty=c(2), lwd=2, bty='n')
  })
  with(pars3, {
	     lines(x.vals, TypeIII(x.vals, beta0, beta1, k), lty=3, lwd=2)
       legend(10, 10, paste0("beta0=", beta0, ", beta1=", beta1, ", k=", k), lty=c(3), lwd=2, bty='n')
  })
}
```
A slightly more general, 3-parameter Type III functional response takes the form:

$$
  c = \frac{\beta_0 x^k}{1 + \beta_0 \beta_1 x^k}
$$

Where the kill rate saturates at $1/\beta_1$, and the additional parameter $k$ determines how quickly (in terms of increasing density) shooting saturation is reached. Higher rates of $k$ enable a more pronounced sigmoidal shape. Note that once prey density increases beyond the minimum (effectively $c_{max}$) the slope should not exceed one. That is, the kill rate shouldn't increase more quickly than the increase in prey density [Specifying statistical priors to achieve this may be a touch challenging ...]. 

```{r, fig.cap="Example of generalised Type III functional response."}
if(PLOT.IT) {
  pars1 <- list(beta0=0.1, beta1=1/40, k=4, x.vals=seq(0, 20, 0.1))
  pars2 <- list(beta0=0.1, beta1=1/40, k=5, x.vals=seq(0, 20, 0.1))
  pars3 <- list(beta0=0.5, beta1=1/40, k=4, x.vals=seq(0, 20, 0.1))
  pars4 <- list(beta0=1.0, beta1=1/40, k=4, x.vals=seq(0, 20, 0.1))
  with(pars1, {
	     plot(x.vals, TypeIIIGeneral(x.vals, beta0, beta1, k), type='l', lwd=2, xlim=c(0,max(x.vals)), xlab="Prey density", ylab="Kill rate", axes=F)
      axis(1, lwd=3, pos=-0.2)
      axis(2, lwd=3, pos=-0.25, las=2)
      abline(h=40, lty=2)
       #lines(x.vals, TypeIIIGeneral(x.vals, beta0=0.05, beta1=0.1, k=4), lty=2)
       legend(10, 30, paste0("beta0=", beta0, ", beta1=", beta1, ", k=", k), lty=c(1), lwd=2, title="Parameter settings", bty='n')
  })
  with(pars3, {
	     lines(x.vals, TypeIIIGeneral(x.vals, beta0, beta1, k), type='l', lwd=2, lty=3, xlim=c(0,max(x.vals)), xlab="Prey density", ylab="Kill rate")
       legend(10, 20, paste0("beta0=", beta0, ", beta1=", beta1, ", k=", k), lty=c(3), lwd=2, bty='n')
  })
  with(pars4, {
	     lines(x.vals, TypeIIIGeneral(x.vals, beta0, beta1, k), type='l', lwd=2, lty=4, xlim=c(0,max(x.vals)), xlab="Prey density", ylab="Kill rate")
       legend(10, 15, paste0("beta0=", beta0, ", beta1=", beta1, ", k=", k), lty=c(4), lwd=2, bty='n')
  })
}
```

## Choice of functional response

Do we need to species-dependent choices, or species x habitat-dependent choices, or can a 3-parameter Type III cover all? 


# Fitting functional response models

Fitting functional response models requires that density is known (from a pre-control survey) or can be estimated jointly. Analyses to date [e.g. @Choquenot.etal:1999] have tended to assume the pre-control population is measured without uncertainty. A better approach would be to include uncertainty in the estimate of the pre-control population density. This naturally sends us down the Bayesian path.

```{r}
# Fitting functional response. 



```


# Simulation of shooting data from known populations

It will be useful for model testing and development to be able to simulate realistic removal data from a known population.


```{r, echo=FALSE}
simShoot <- function(n0=40, A=10, T=10, c.max=40, x.min=2, d=0.5, beta1=20, k=2, effort=NULL, effort.mu=6, searchFunc="Ivlev") {
# Args:
#   n0 -- initial population density
#   N0 -- initial population size  
#   A -- area inhabited
#   T -- number of shooting events
#   c.max -- maximum off-take rate per unit area from Ivlev 
#   x.min -- density below which animals aren't available
#   d -- rate of amelioration of off-take rate with decreasing density
#   effort --- vector of shooting effort
#   effort.mu -- mean number of hours of per shoot (if effort not specified)
# Returns: 
#   A list 
  x <- rep(NA, T)
  cull.prop <- rep(0, T)
  cull.no <- rep(0, T)
  cull.rate <- rep(NA, T)
  mu <- rep(NA, T)
  if(is.null(effort)) {effort <- rep(effort.mu, T)}
  N <- rep(NA, T)
  N[1] <- n0*A
  # Density initially
  x[1] <- n0
  # Kill encounter rate per unit time
  if(searchFunc=="Ivlev") {
      mu[1] <-  Ivlev(x[1], c.max=c.max, x.min=x.min, d=d)}
  if(searchFunc=="Choqu") {
       mu[1] <-  Choquenot(x[1], a=c.max, b=x.min, d=d)}
  if(searchFunc=="TypeIII"){
       beta0 <- c.max 
       mu[1] <- TypeIII(x[1], beta0=beta0, beta1, k)}
  # Add variation
  # mu[1] <- rpois(1, mu[1])
  # ToDo -- check below here ... 
  # Encounter & probability per capita
  # cull.prop[1] <- 1 - exp(-mu[1]/(A*x[1])*effort[1])
  # Number culled with standard binomial variation
  # cull.no[1] <- rbinom(1, N[1], cull.prop[1])
  # Number culled per unit effort
  # cull.rate[1] <- cull.no[1]/effort[1]
  cull.no[1] <- rpois(1, mu[1]*effort[1])
  # Run simulation
  for(i in 2:T) { # i=1
     # Number remaining
     N[i] <- N[i-1] - cull.no[i-1]
     # Update density alive
     x[i] <- N[i]/A
     # Kill encounter rate per unit time
     if(searchFunc=="Ivlev") {
      mu[i] <-  Ivlev(x[i], c.max=c.max, x.min=x.min, d=d)}
     if(searchFunc=="Choqu") {
       mu[i] <-  Choquenot(x[i], a=c.max, b=x.min, d=d)}
     if(searchFunc=="TypeIII"){
       beta0 <- c.max 
       mu[i] <- TypeIII(x[i], beta0=beta0, beta1, k)
     }
     # Add variation
     # mu[i] <- rpois(1, mu[i])
     # Encounter & probability per capita
     # cull.prop[i] <- 1 - exp(-mu[i]/(A*x[i])*effort[i])
     # Number culled with standard binomial variation
     # cull.no[i] <- rbinom(1, N[i], cull.prop[i])
     # Number culled per unit effort
     # cull.rate[i] <- cull.no[i]/effort[i]
     cull.no[i] <- rpois(1, mu[i]*effort[i])  
  }
  return(list(
    mu=mu,
    cull.prop=cull.prop,
    cull.no=cull.no,
    cull.rate=cull.rate,
    cull.eff=effort,
    effort.mu=effort.mu,
    x=x,
    N=N,
    t=1:T,
    x.min=x.min,
    n0=n0,
    A=A,
    c.max=c.max,
    x.min=x.min,
    d=d,
    beta0=c.max,
    beta1=beta1,
    k=k,
    T=T
    ))
}

if(TEST) {
  simShoot(n0=20, A=10, T=10, c.max=10, x.min=1.5, d=0.5, effort.mu=6, effort=NULL, searchFunc="Ivlev" )
  simShoot(n0=20, A=10, T=10, c.max=10, x.min=1.5, d=0.5, effort.mu=6, effort=NULL, searchFunc="Choqu" )
}
```

```{r, echo=TRUE, fig.cap="(A) Simulated removals, and (B) Simulated culling rate assuming an Ivlev functional response."}
# Ivlev example 
out <- simShoot(n0=20, A=100, T=10, c.max=40, x.min=2, d=0.5, effort.mu=6, effort=NULL, searchFunc="Ivlev" )
# Plot it
with(out, {
  par(mfrow=c(1,2))
  plot(1:T, cull.no, type='h', lwd=2, ylab="Removals", xlab="Shooting session", axes=FALSE)
  axis(1, lwd=2, at=t)
  axis(2, lwd=2, las=2)
  mtext("(A)", side=3, adj=0, cex=1.5)
  plot(1:T, mu, pch=16, cex=2, lwd=2, ylab="Underlying removal rate (/hour)", xlab="Shooting session", axes=FALSE, ylim=c(0, max(mu, na.rm=TRUE)))
  axis(1, lwd=2, at=t)
  axis(2, lwd=2, las=2)
  mtext("(B)", side=3, adj=0, cex=1.5)
  legend(1, 20, paste0("c.max=", c.max, ", x.min=", x.min, ", d=", d), lty=(1), lwd=2, title="Shooting parameter settings")
  legend(1, 10, paste0("Inital density=", n0), lty=(1), lwd=2, title="Population settings")
  })
```

Generate some simulated data.

```{r, fig.cap="(A) Simulated removals, and (B) Simulated culling rate assuming a Choquenot-like response.", include=TRUE, echo=FALSE}
# Choquenot example
out <- simShoot(n0=20, A=100, T=10, c.max=50, x.min=2, d=0.5, effort.mu=6, effort=NULL, searchFunc="Choqu" )
# Plot it
with(out, {
  par(mfrow=c(1,2))
  plot(t, cull.no, type="h", lwd=2, ylim=c(0, max(cull.no)), ylab="Removals", xlab="Shooting session", axes=FALSE)
  axis(1, lwd=2, at=t)
  axis(2, lwd=2, las=2)
  mtext("(A)", side=3, adj=0, cex=1.5)
  plot(t, mu, pch=16, cex=2, lwd=2, ylim=c(0, max(mu, na.rm=TRUE)), ylab="Underlying mean removals (h-1)", xlab="Shooting session", axes=FALSE)
  axis(1, lwd=2, at=t)
  axis(2, lwd=2, las=2)
  mtext("(B)", side=3, adj=0, cex=1.5)
  legend(1, 20, paste0("c.max=", c.max, ", x.min=", x.min, ", d=", d), lty=(1), lwd=2, title="Shooting parameter settings")
  legend(1, 10, paste0("Inital density=", n0), lty=(1), lwd=2, title="Population settings")
  }
)
```

```{r, fig.cap="(A) Simulated removals, and (B) Simulated culling rate assuming a Type III functional functional response.", include=TRUE, echo=FALSE}
# TypeIII example
simTypeIII <- simShoot(n0=20, A=100, T=10, c.max=50, beta1=20, k=4, effort.mu=6, effort=NULL, searchFunc="TypeIII" )
# Plot it
with(simTypeIII, {
  par(mfrow=c(1,2))
  plot(t, cull.no, type="h", lwd=2, ylim=c(0, max(cull.no)), ylab="Removals", xlab="Shooting session", axes=FALSE)
  axis(1, lwd=2, at=t)
  axis(2, lwd=2, las=2)
  mtext("(A)", side=3, adj=0, cex=1.5)
  plot(t, mu, pch=16, cex=2, lwd=2, ylim=c(0, max(mu, na.rm=TRUE)), ylab="Underlying mean removals (h-1)", xlab="Shooting session", axes=FALSE)
  axis(1, lwd=2, at=t)
  axis(2, lwd=2, las=2)
  mtext("(B)", side=3, adj=0, cex=1.5)
  legend(1, 20, paste0("c.max=", c.max, ", beta0=", beta0), lty=(1), lwd=2, title="Shooting parameter settings")
  legend(1, 10, paste0("Inital density=", n0), lty=(1), lwd=2, title="Population settings")
  }
)
```


# Removal estimators 

Contexts for the use of removal estimators include:

- single estimate of culls per unit area (i.e. no area re-searched in discrete session)
- ...



## Approach of Ramsey *et al.* (2023)

@Ramsey.etal:2023 don't choose a particular functional response when developing a removal estimator for deer populations subject to repeated periods of shooting, where each period comprises several shooting sessions. Key differences to the previous papers discussed include the absence of density dependence in the probability of being shot (i.e. no saturation in the culling rate arising from a fixed handling time per animal). 


## Approaches using parametric forms for the functional response

First up a model using a Type III numerical response. 

```{r, eval=TRUE, echo=TRUE, include=!FALSE}
# Model for Type III estimation
cat("
model {
  # Priors
  n0 ~ dnorm(n0.hat, pow(se.n0.hat,-2))T(0,)
  beta0 ~ dgamma(beta0_shape, beta0_rate) 
  beta1 ~ dgamma(beta1_shape, beta1_rate) 
  k ~ dgamma(k_shape, k_rate)         
  
  N0 <- n0*A
  N[1] <- round(N0)
  # Density
  n[1] <- n0 
  # Kill rate (Type III)
  # Functional Response model for cull rate
  mu[1] <- beta0*n[1]^k/(beta1^2 + n[1]^k)
  # Culls likelihood
  x[1] ~ dpois(mu[1]*cull.eff[1])
  # Over shooting sessions
  for (i in 2:max(T)) {
    # Remove culls at end of previous period
    N[i] <- N[i-1] - x[i-1]   
    # Calculate density
    n[i] <- N[i]/A
    # Functional Response model for cull rate
    mu[i] <- beta0*n[i]^k/(beta1^2 + n[i]^k)
    # Culls likelihood
    x[i] ~ dpois(mu[i]*cull.eff[i])
  } 
   
  # Predictive distribution
  for(j in 1:length(n_pred)) {
    mu_pred[j] <- beta0*n_pred[j]^k/(beta1^2 + n_pred[j]^k)
  }
}", file="estTypeIII.jags")
```


### Recovering parameters from known simulation ... 

First up we examine whether we can recover the latent (unobserved) states  of the populations underlying simulated shooting datasets. 

```{r, echo=FALSE, include=FALSE}
# Data
mod.data.sim <- list(T=simTypeIII$t, 
                n0.hat=20,
                se.n0.hat=10, 
                beta0_shape=5,  beta0_rate=1/10,# mean 50
                beta1_shape=5, beta1_rate=1/4,   # mean 20
                k_shape=2, k_rate=1/2, 
                cull.eff=simTypeIII$cull.eff,
                x=simTypeIII$cull.no,
                A=simTypeIII$A,
                c.max=simTypeIII$c.max,
                x.min=simTypeIII$x.min,
                d=simTypeIII$d,
                n_pred=seq(0, simTypeIII$n0, 0.5))

# Initialize model
estTypeIII.sim <- jags.model(
  file="estTypeIII.jags",
  dat=mod.data.sim,
  n.chains=3,
  n.adapt=1000 
)

# Burn in
update(estTypeIII.sim, 1E3)

# Parameters of interest
params <- c("N0","N", "n", "k", "beta0", "beta1", "mu", "mu_pred", "p.cull")

# Get samples
estTypeIII.sim.coda <- coda.samples(estTypeIII.sim, params, 1E3) 
# Combine into posterior
post <- as.matrix(estTypeIII.sim.coda)
# Have a look
head(post)
```


Check out posteriors. 


```{r, echo=FALSE}
par(mfrow=c(1,3))
#hist(post[,'N[1]'])
#hist(post[,'n[1]'], main="", xlab="Initial population density")
hist(post[,"N0"], main="", xlab="Initial population size")
abline(v=simTypeIII$n0*simTypeIII$A, lwd=2, col='red')
hist(post[,'beta0'], main="", xlab="Maxiumum kill rate", prob=TRUE)
hist(post[,'k'], main="", xlab="k", prob=TRUE)
```

Check stuff ... 


```{r, fig.cap="ToDo"}
# Calculate mean kill rates 
mu.col.names <- paste0("mu[",simTypeIII$t,"]")
mu.means <- apply(post[ , mu.col.names],2,mean)
mu.pred.col.names <- paste0("mu_pred[", 1:length(mod.data.sim$n_pred), "]")
mu.pred.means <- apply(post[, mu.pred.col.names], 2, mean)
n.col.names <- paste0("n[", simTypeIII$t, "]")
n.means <-  apply(post[,n.col.names], 2, mean)

with(simTypeIII,{
  plot(x, cull.no/cull.eff, pch=16, col='blue', xlim=c(0,n0), ylim=c(0,max(1.2*simTypeIII$beta0)),
       xlab="Estimated population density", ylab="Kill rate")
})
lines(mod.data.sim$n_pred, mu.pred.means, lwd=2)

```


#  Fitting different models to existing datasets

## Feral/wild pigs

@Hone:1990PredPrey and @Saunders&Bryant:1988 reported on aerial shooting operation operations of feral pigs on the Mary River floodplain (Northern Territory) and Macquarie Marshes (western NSW), respectively.  

```{r, echo=FALSE}
# Data from Hone 1990 from aerial shooting operation on the Mary River floodplain. 
MR1986 <- data.frame(
  Day=1:9,
  EffortKm2=c(0.015, 0.022, 0.016, 0.006, 0.016, 0.018, 0.007, 0.019, 0.016),
  KillsKm2=c(1.017, 0.471, 0.627, 0.210, 0.892, 0.566, 0.414, 0.519, 0.146),
  KillRate=c(67.8, 21.4, 39.2, 35.0, 55.8, 31.4, 59.1, 27.3, 9.1)
)
# Calculate kills over entire site
MR1986 <- within(MR1986, {
    Kills <- round(KillsKm2*295)
    Effort <- EffortKm2*295
})
MR1986 <- MR1986[ ,c("Day", "Effort", "Kills", "KillRate", "KillsKm2")]

# Data from Saunders & Bryant (1988) for the Macquarie Marshes in 1985
MM1985 <- data.frame(
  Day=1:4,
  Effort=c(255,405,420,330)/60,
  Kills=c(208,417,316,74)
)
MM1985 <- within(MM1985, {
  KillsKm2 <- round(Kills/120, 2)
  KillRate <- round(Kills/Effort, 2)
})
# Make table
kable(MR1986, caption="Aerial shooting data of feral pigs from Mary River Flooplain, Northern Territory. Reproduced from @Hone:1990PredPrey.")

kable(MM1985, caption="Aerial shooting data of feral pigs from Macquarie Marshes, Western NSW. Reproduced from @Saunders&Bryant:1988.")
```

A notable feature of the Mary River operation data is the modest decline in the kill rate (see Figure below). This may be a result of the shooters  pseudo-systematic search pattern, whereby "shooting began on the perimeter of the area and continued towards the centre." Hence some of the kill rates for later shooting  sessions (e.g. Day 7) could arise from areas being shot for the first time. Other considerations for model specification include most of the open floodplain being covered with 1--2 metres of water. If the approx. one third of the study areas that was floodplain was subtracted from the total (295 km$^2$) this would change the estimates. 

```{r, fig.cap="Data from aerial shooting operation reported by (A) @Hone:1990PredPrey and (B) @Saunders&Bryant:1988.", echo=FALSE}
par(mfrow=c(1,2))
with(MR1986,
     plot(Day, KillRate, pch=16, ylim=c(0, max(KillRate)), axes=FALSE, 
          cex=1.5,
          ylab=expression(Kill ~ rate ~ hr^-1)))
  axis(1, lwd=2, at=1:9)
  axis(2, lwd=2, las=2)
  mtext("(A)", 3, adj=0)

with(MM1985,
     plot(Day, KillRate, pch=16, ylim=c(0, max(KillRate)), axes=FALSE, 
          cex=1.5,
          ylab=expression(Kill ~ rate ~ hr^-1)))
  axis(1, lwd=2, at=1:9)
  axis(2, lwd=2, las=2)
  mtext("(B)", 3, adj=0)
  
```


### Fitting Macquarie Marshes feral pig shoot data of @Saunders&Bryant:1988.

<<<<<<< HEAD
The details in @Saunders&Bryant:1988 of the pre-shoot aerial survey (where ?? animals were counted) precludes estimating population density, so instead a prior of 50 pigs km$^-2$ was used consistent with some of the previous data from Jack Giles. Shooting was undertaken using a Bell 206 Jetranger helicopter. 
=======
ToDo. The details in @Saunders&Bryant:1988 of the pre-shoot aerial survey (where ?? animals were counted) precludes estimating population density, so instead a prior of 50 pigs km$^-2$ was used consistent with some of the previous data from Jack Giles. 
>>>>>>> f60a3e6ab6da29bea1dabe2bac7f2c4f989fff3b

```{r, echo=FALSE, include=FALSE}
mod.data.MM1985 <- list(T=1:4, 
                n0.hat=50,   # from Giles (1980)
                se.n0.hat=0.5*50, # Placeholder ... pretty flat
                beta0_shape=5,  beta0_rate=1/10,# mean 50
                beta1_shape=5, beta1_rate=1/4,   # mean 20
                k_shape=1, k_rate=1/10,
                cull.eff=MM1985$Effort,
                cull.rate=MM1985$KillRate,
                x=MM1985$Kills,
                A=120,
                n_pred=seq(0,20,0.5)
                )

# Check out data

# Initialize model
simShoot.MM1985 <- jags.model(
  file="estTypeIII.jags",
  dat=mod.data.MM1985,
  n.chains=3,
  n.adapt=1000 
)

# Burn in
update(simShoot.MM1985, 1E3)

# Parameters of interest
params <- c("N0","N", "n", "mu", "mu_pred", "k", "p.cull")

# Get samples
simShoot.MM1985.coda <- coda.samples(simShoot.MM1985, params, 1E3) 

post <- as.matrix(simShoot.MM1985.coda)

head(post)
```

```{r, fig.cap="Posterior distributions (histograms) with priors (green lines) for (A) Initial population density, (B) Initial population density on wider scale to better illustrate prior, and (C) Parameter $k$."}
par(mfrow=c(1,3))
#hist(post[,'N[1]'], main="Initial population size", probability=TRUE)
hist(post[,'n[1]'], main="", probability=TRUE, xlab="Initial population density")
curve(dnorm(x, mean=50, sd=0.5*50), add=TRUE, col='green', lwd=2, )
mtext("(A)", 3, adj=0)
hist(post[,'n[1]'], main="", probability=TRUE, xlab="Initial population density", xlim=c(0,100))
# add prior
curve(dnorm(x, mean=50, sd=0.5*50), add=TRUE, col='green', lwd=2, )
mtext("(B)", 3, adj=0)
hist(post[,'k'], main="", probability=TRUE, xlab="k")
curve(dgamma(x, mod.data.MM1985$k_shape, mod.data.MM1985$k_rate), add=TRUE, col='green', lwd=2)
mtext("(C)", 3, adj=0)
```

```{r, fig.cap="Kill rates as a function of estimated density for feral pigs on the Maquaries Marshes, western NSW. Source: @Saunders&Bryant:1988"}
# Calculate mean kill rates 
mu.col.names <- paste0("mu[", MM1985$Day, "]")
mu.means <-  apply(post[, mu.col.names], 2, mean)
mu.pred.col.names <- paste0("mu_pred[", 1:length(mod.data.MM1985$n_pred), "]")
mu.pred.means <- apply(post[, mu.pred.col.names], 2, mean)
n_hat.col.names <- paste0("n[", MM1985$Day, "]")
n_hat.means <-  apply(post[,n_hat.col.names], 2, mean)

with(mod.data.MM1985, {
     plot(n_hat.means, cull.rate, pch=16, col='blue', ylim=c(0, 1.2*max(cull.rate)), axes=FALSE, xlim=c(0, max(n_hat.means)),
          xlab="Estimated density", ylab="Cull rate")
  axis(1, lwd=3)
  axis(2, lwd=2, las=2)
})
lines(mod.data.MM1985$n_pred, mu.pred.means, pch=16, ylim=c(0,300), lwd=2, 
      xlab="Estimated density", ylab="Kill rate")

```



### Mary River floodplain shoot from @Hone:1990PredPrey

This shoot was undertaken in the wet season, with much of the floodplain covered in deep water. Shooting was undertaken from a Bell 47 helicopter.

```{r, echo=FALSE, include=FALSE}
mod.data.MR1986 <- list(T=1:nrow(MR1986), 
                #n0.hat=50,   # from Giles (1980)
                #se.n0.hat=0.5*50, # Placeholder ... pretty flat
                n0.hat=6.1,   # from Hone (1990)
                se.n0.hat=0.5, # ToDo Guess from Hone(1990)
                beta0_shape=5,  beta0_rate=1/10, # mean 50
                beta1_shape=1, beta1_rate=1/10,   # mean 20
                k_shape=1, k_rate=1/10,
                cull.eff=MR1986$Effort,
                cull.rate=MR1986$KillRate,
                x=MR1986$Kills,
                n_pred=seq(0,10,0.1),
                A=295
                )

# Initialize model
simShoot.MR1986 <- jags.model(
  file="estTypeIII.jags",
  dat=mod.data.MR1986,
  n.chains=3,
  n.adapt=1000 
)

# Burn in
update(simShoot.MR1986, 1E3)

# Parameters of interest
params <- c("N0","N", "n", "beta0", "mu", "mu_pred", "k", "p.cull")

# Get samples
simShoot.MR1986.coda <- coda.samples(simShoot.MR1986, params, 10E3) 

post <- as.matrix(simShoot.MR1986.coda)

head(post)
```


```{r, echo=FALSE, fig.cap="Posterior estimates for fitting model to shooting data of feral pigs from the Mary River Floodplain, Northern Territory."}
par(mfrow=c(2,2))
#hist(post[,'N[1]'], main= "", probability=TRUE, xlab="Initial population size")
hist(post[,'n[1]'], main="", probability=TRUE, xlab="Initial population density")
abline(v=6.13, lwd=2, col='red')
hist(post[,'n[1]'], main="", probability=TRUE, xlim=c(0,100), xlab="Initial population density")
# add prior
curve(dnorm(x, mean=50, sd=0.25*50), add=TRUE)
hist(post[,'beta0'], main="", xlab="Maximum kill rate")
hist(post[,'k'], main="", xlab="k")
```


```{r, fig.cap="Kill rates as a function of estimated density for feral pigs on the Mary River Floodplain, Northern Territory. Source: @Hone:1990PredPrey", echo=FALSE}
# Calculate mean kill rates 
mu.col.names <- paste0("mu[", MR1986$Day, "]")
mu.means <-  apply(post[, mu.col.names], 2, mean)
mu.pred.col.names <- paste0("mu_pred[", 1:length(mod.data.MR1986$n_pred), "]")
mu.pred.means <- apply(post[, mu.pred.col.names], 2, mean)
n_hat.col.names <- paste0("n[", MR1986$Day, "]")
n_hat.means <-  apply(post[,n_hat.col.names], 2, mean)

with(mod.data.MR1986, {
     plot(n_hat.means, cull.rate, pch=16, col='blue', ylim=c(0, max(cull.rate)), axes=FALSE, xlim=c(0,max(n_hat.means)),
          xlab="Estimated density", ylab="Kill rate")
  axis(1, lwd=3)
  axis(2, lwd=2, las=2)
})
lines(mod.data.MR1986$n_pred, mu.pred.means, pch=16, ylim=c(0,300), lwd=2)
```
### Combining multiple datasets ...
```{r}
mod.data.comb <- list(
                T=c(nrow(MM1985), nrow(MR1986)), 
                n0.hat=c(50, 50),   # from Giles (1980)
                se.n0.hat=c(0.5*50, 0.5*50), # Placeholder ... pretty flat
                beta0_shape=5,  beta0_rate=1/10, # mean 50
                beta1_shape=1, beta1_rate=1/10,   # mean 20
                k_shape=1, k_rate=1/10,
                cull.eff=list(MM1985$Effort, MR1986$Effort),
                cull.rate=MR1986$KillRate,
                cull.no=MR1986$Kills,
                x_pred=seq(0,10,0.1),
                A=295
                )
```


```{r, eval=TRUE, echo=TRUE, include=!FALSE}
# Model for Type III estimation with multiple data sets 
cat("
model {
  # Priors
  for (j in 1:J){
    n0[j] ~ dnorm(n0.hat[j], pow(se.n0.hat[j],-2))T(0,)
  }
  beta0 ~ dgamma(beta0_shape, beta0_rate) 
  beta1 ~ dgamma(beta1_shape, beta1_rate) 
  k ~ dgamma(k_shape, k_rate)         
  # Across sites
  for (j in 1:J) {
    N0[j] <- n0[j]*A[j]
    N[1,j] <- round(N0[j])
    # Density
    n[1,j] <- n0[j] 
    # Functional Response model for cull rate (Type III)
    mu[1,j] <- beta0*n[1,j]^k/(beta1^2 + n[1,j]^k)
    # Culls likelihood
    cull.no[1,j] ~ dpois(mu[1,j]*cull.eff[1,j])
    # Over shooting sessions
    for (i in 2:T[j]) {
      # Remove culls at end of previous period
      N[i,j] <- N[i-1,j] - x[i-1,j]   
      # Calculate density
      n[i,j] <- N[i,j]/A[j]
      # Functional Response model for cull rate
      mu[i,j] <- beta0*n[i,j]^k/(beta1^2 + n[i,j]^k)
      # Culls likelihood
      x[i,j] ~ dpois(mu[i,j]*cull.eff[i,j])
    } 
  } 
  # Predictive distribution for cull rate
  for(m in 1:length(n_pred)) {
    mu_pred[m] <- beta0*n_pred[m]^k/(beta1^2 + n_pred[m]^k)
  }
}", file="estTypeIIIcomb.jags")
```


```{r}
# Add missing data to Macquarie Marshes shoot
extra <- as.data.frame(cbind(5:9, matrix(NA, nrow=5, ncol=4)))
names(extra) <- names(MM1985)
MM1985PadNA <- rbind(MM1985, extra)

data.comb <- list(
                J=2, # Number of sites
                T=c(max(MM1985$Day), max(MR1986$Day)), # number of shooting sessions
                n0.hat=c(50, 6.1),   # from Giles (1980)
                #se.n0.hat=c(0.5*50, 0.5), # Placeholder ... pretty flat
                se.n0.hat=c(0.5*50, 5*6.1), # Placeholder ... pretty flat
                beta0_shape=5,  beta0_rate=1/10, # mean 50
                beta1_shape=1, beta1_rate=1/10,   # mean 20
                k_shape=1, k_rate=1/10,
                cull.eff=cbind(MM1985PadNA$Effort, MR1986$Effort),
                cull.rate=cbind(MM1985PadNA$KillRate, MR1986$KillRate),
                x=cbind(MM1985PadNA$Kills, MR1986$Kills),
                n_pred=seq(0,10,0.1),
                A=c(120,295)
                )

# Initialize model
simShoot.comb <- jags.model(
  file="estTypeIIIComb.jags",
  dat=data.comb,
  n.chains=3,
  n.adapt=1000 
)

# Burn in
update(simShoot.comb, 1E3)

# Parameters of interest
params <- c("n0", "N0","N", "n", "beta0", "beta1", "mu", "mu_pred", "k")

# Get samples
simShoot.comb.coda <- coda.samples(simShoot.comb, params, 10E3) 

post.comb <- as.matrix(simShoot.comb.coda)

#head(post.comb)

```


```{r, fig.cap="Kill rates of feral pigs during aerial shooting operations on the Mary River floodplain and Macquarie Marshes. Model fitted jointly to datasets with same parameterisation. ", echo=FALSE}
# Calculate mean kill rates 
MM.mu.col.names <- paste0("mu[", paste(1:4,1,sep=","), "]")
MM.mu.means <-  apply(post.comb[, MM.mu.col.names], 2, mean)

MR.mu.col.names <- paste0("mu[", paste(1:9,2,sep=","), "]")
MR.mu.means <-  apply(post.comb[, MR.mu.col.names], 2, mean)

MM.n_hat.col.names <- paste0("n[", paste(1:4,1,sep=","), "]")
MM.n_hat.means <-  apply(post.comb[ ,MM.n_hat.col.names], 2, mean)

MR.n_hat.col.names <- paste0("n[", paste(1:9,2,sep=","), "]")
MR.n_hat.means <-  apply(post.comb[ ,MR.n_hat.col.names], 2, mean)

mu.pred.col.names <- paste0("mu_pred[", 1:length(data.comb$n_pred), "]")
mu.pred.means <- apply(post.comb[, mu.pred.col.names], 2, mean)

with(data.comb, {
     plot(MM.n_hat.means[1:4], cull.rate[1:4 ,1], pch=16, col='blue', ylim=c(0, max(cull.rate,na.rm=TRUE)), axes=FALSE, xlim=c(0, ceiling(max(MM.n_hat.means, na.rm=TRUE))),
     xlab="Estimated density", ylab="Kill rate")
     points(MR.n_hat.means, cull.rate[ ,2])
  axis(1, lwd=3)
  axis(2, lwd=2, las=2)
})
lines(data.comb$n_pred, mu.pred.means, pch=16, ylim=c(0,300), lwd=2)
legend(6,40,c("Macquarie Marshes 1985", "Mary River Floodplain 1986"), pch=c(16,1), col=c("blue",1), title="Shooting operation")
```

There is clearly something interesting going on here. 

```{r}
par(mfrow=c(3,3))
hist(post.comb[,"beta1"])
plot(post.comb[,"k"], post.comb[,"n0[2]"])
plot(post.comb[,"beta0"], post.comb[,"beta1"])
plot(post.comb[,"beta0"], post.comb[,"n0[2]"])
plot(post.comb[,"beta1"], post.comb[,"n0[2]"])
```



## Wild deer

Re-fit data from @Bengsen.etal:2023. 



# References



